Copulas and their potential for ecology

### 2019-12-17 Paper.Rmd
fig in greyscale fig_s25-s29, fig_15

We approached our overall goal of exploring the potential of copula approaches 
for ecology by addressing the following specific questions. (Q1) Do datasets
in ecology have dependence structure distinct from that of a multivariate 
Gaussian/normal
distribution (here called non-normal copula structure)? Do positvely associated 
ecological variables show tail 
associations distinct from those of a normal distribution, and in particular
do they show asymmetric tail associations? Normal copula structure is assumed
by standard approaches that use multivariate 
normal distributions or distributions obtained by transforming
the marginals of a normal distribution. So Q1 asks whether ecological
data contain dependency information distinct from a standard or default case. 
(Q2) If the answer to
Q1 is "yes," then what are some possible causes/mechanisms of non-normal
copula structure and asymmetric tail associations in ecology? And (Q3), what are the 
consequences of non-normal copula
structure and asymmetric tail associations for ecological understanding and
applications? After section \ref{Background}, 'Background on copulas', we address Q1 
(sections \ref{Data} through \ref{Results} of the paper) 
by analyzing several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data, selected to span multiple 
sub-fields and organizational levels of ecology. 
We address Q2 (sections \ref{Causes} through \ref{Results_Q2} of the paper), using 
simple models. We address Q3 (sections \ref{Consequences} through \ref{Results_Q3} 
of the paper) using 
both data and models. Section \ref{Discussion} is the Discussion.
Multiple analyses are brought to bear on each question, and these
are summarized diagramatically in Fig. \ref{MasterFigure}, which can 
serve as a *post hoc* guide to the paper. 
Copula approaches have been used to great effect in neuroscience [@onken2009], 
bioinformatics [@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], and finance [@Li2000],
and have also been used very effectively, but rarely so far, in ecology [@Valpine2014; 
@Anderson2018; @Popovic2018]. 
We argue that benefits of wider usage of complete 
copula descriptions of dependence in ecology will be substantial.

@@

Recall that Q1 has three parts: Do datasets
in ecology have dependence structure distinct from that of a multivariate 
Gaussian/normal
distribution (here called non-normal copula structure)? Do positvely associated 
ecological variables show tail 
associations distinct from those of a normal distribution? And in particular
do they show asymmetric tail associations? 
We addressed Q1 first via a model selection procedure in which several
families of copulas were fitted to 
our ecological datasets and fits were compared via the Akaike and Bayesian 
Information Criteria (AIC and BIC). One of the fitted copulas was the normal
copula, making possible comparisons of the degree to which the normal copula versus
other copula families were good descriptions of data.

@@

<!--Plots for AR(1) with Clayton-->
\begin{figure}[!h]
\begin{center}
\includegraphics[width=13 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/common_legend_cause4copula_stat.pdf}\\
\vspace{-0.4 cm}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Spearman_vs_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Kendall_vs_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Pearson_vs_Kendall's Tau.pdf}\\
\vspace{-0.2 cm}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Corl_vs_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Coru_vs_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Corl-Coru_vs_Kendall's Tau.pdf}\\
\vspace{-0.2 cm}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Pl_vs_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Pu_vs_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Pl-Pu_vs_Kendall's Tau.pdf}\\
\vspace{-0.2 cm}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_D2u_vs_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_D2l_vs_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_D2u-D2l_vs_Kendall's Tau.pdf} \\
\hspace{-0.6 cm}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_scatter_CorlmCoru_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_scatter_PlmPu_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_scatter_D2umD2l_Kendall's Tau.pdf}
\caption{Comparison of correlation and tail association statistics between 
environmental-noise inputs and ecological-variable 
outputs, AR(1) model, Clayton copula. Red and blue points give 
means over 50 replicate simulations of the listed statistics, 
red points are for noise inputs and blue points are for
model outputs. Error bars are standard errors. 
$P$-values (purple points, right axis) are for a 
paired $t$-test of the null hypothesis that the distributions have the same mean. 
For bottom panels, headers give Pearson correlations for
the points. The regression line through the points (black) was similar
to the 1-1 line (green). 
See section \ref{Causes} for details.\label{fig_Cause4copula_C_kend_ar1}}
\end{center}
\end{figure}

@@

<!--adding plots showing influences of non-normal copula on spatial Taylors law-->
\begin{figure}[!h]
\textbf{ \hspace{2cm} (A) \hspace{2.92 cm} (B) \hspace{2.92 cm} (C) \hspace{2.92 cm} (D)} \\
\vspace{-0.8 cm} 
\begin{center}
\includegraphics[width=16 cm]{./Results/taylorslaw_results/SpatialTL_CNFSC_spearman.png}\\
\end{center}
\textbf{ \hspace{1 cm} (E) \hspace{1.4 cm} (F) \hspace{1.2 cm} (G) \hspace{1.2 cm} (H) \hspace{1.4 cm} (I) \hspace{1.4 cm} (J) \hspace{1.2 cm} (K)}\\
\vspace{-0.8 cm}
\begin{center}
\includegraphics[width=16 cm]{./Results/taylorslaw_results/SpatialTL_7stats_spearman.pdf}
\caption[Spatial Taylor's law results]{Spatial Taylor's law results. Log(spatial variance) vs. 
log(spatial mean) relationships for $1000$ simulations over $50$ time steps for $25$ locations 
using Clayton (A), normal (B), Frank (C), and survival Clayton (D) copula structures between
locations. Results from all $1000$ 
simulations were plotted on the same axes for each panel, but two example simulations are 
shown in red and green to help assess to what extent variation on the plots was between 
simulations or within simulations. For each simulation, we tested the linearity 
(E) and homoskedasticity (F) of the log(variance) vs. log(mean) plot for that simulation, 
and quantified the root mean squared error of points from the linear regression line (G).
We also calculated the linear regression intercept (H) and slope (I), the quadratic
term of a quadratic regression through the points on the log(variance) vs. log(mean) plot
(J), and the curvature of that quadratic regression (K). Distributions of values 
across all 1000 simulations are displayed. Spearman-preserving surrogates were used, 
though results using Kendall-preserving surrogates were similar. See section \ref{Consequences}
and Appendices \ref{SM-Consequence_TL_simulation} and 
\ref{SM-sect:copsurrognd} for details.\label{fig_spTL}}
\end{center}
\end{figure}

>>>>

We approached our overall goal of exploring the potential of copula approaches 
for ecology by addressing the following specific questions. (Q1) Do datasets
in ecology have dependence structure distinct from that of a multivariate 
Gaussian/normal
distribution (here called non-normal copula structure)? Do positively associated 
ecological variables show tail 
associations distinct from those of a normal distribution, and in particular
do they show asymmetric tail associations? Normal copula structure is assumed
by standard approaches that use multivariate 
normal distributions or distributions obtained by transforming
the marginals of a normal distribution. So Q1 asks whether ecological
data contain dependency information distinct from a standard or default case. 
(Q2) If the answer to
Q1 is "yes," then what are some possible causes/mechanisms of non-normal
copula structure and asymmetric tail associations in ecology? And (Q3), what are the 
consequences of non-normal copula
structure and asymmetric tail associations for ecological understanding and
applications? After section \ref{Background}, 'Background on copulas', we address Q1 
(sections \ref{Data} through \ref{Results} of the paper) 
by analyzing several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data, selected to span multiple 
sub-fields and organizational levels of ecology. 
We address Q2 (sections \ref{Causes} through \ref{Results_Q2} of the paper), using 
simple models. We address Q3 (sections \ref{Consequences} through \ref{Results_Q3} 
of the paper) using 
both data and models. Section \ref{Discussion} is the Discussion.
Multiple analyses are brought to bear on each question, and these
are summarized diagramatically in Fig. \ref{MasterFigure}, which can 
serve as a *post hoc* guide to the paper. 
Copula approaches have been used to great effect in neuroscience [@onken2009], 
bioinformatics [@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], and finance [@Li2000],
and have also been used very effectively, but rarely so far, in ecology [@Valpine2014; 
@Anderson2018; @Popovic2018]. 
We argue that benefits of wider usage of complete 
copula descriptions of dependence in ecology will be substantial.

@@

Recall that Q1 has three parts: Do datasets
in ecology have dependence structure distinct from that of a multivariate 
Gaussian/normal
distribution (here called non-normal copula structure)? Do positively associated 
ecological variables show tail 
associations distinct from those of a normal distribution? And in particular
do they show asymmetric tail associations? 
We addressed Q1 first via a model selection procedure in which several
families of copulas were fitted to 
our ecological datasets and fits were compared via the Akaike and Bayesian 
Information Criteria (AIC and BIC). One of the fitted copulas was the normal
copula, making possible comparisons of the degree to which the normal copula versus
other copula families were good descriptions of data.

@@

<!--Plots for AR(1) with Clayton-->
\begin{figure}[!h]
\begin{center}
\includegraphics[width=13 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/common_legend_cause4copula_stat.pdf}\\
\vspace{-0.4 cm}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Spearman_vs_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Kendall_vs_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Pearson_vs_Kendall's Tau.pdf}\\
\vspace{-0.2 cm}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Corl_vs_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Coru_vs_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Corl-Coru_vs_Kendall's Tau.pdf}\\
\vspace{-0.2 cm}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Pl_vs_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Pu_vs_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Pl-Pu_vs_Kendall's Tau.pdf}\\
\vspace{-0.2 cm}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_D2u_vs_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_D2l_vs_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_D2u-D2l_vs_Kendall's Tau.pdf} \\
\hspace{-0.6 cm}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_scatter_CorlmCoru_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_scatter_PlmPu_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_scatter_D2umD2l_Kendall's Tau.pdf}
\caption{Comparison of correlation and tail association statistics between 
environmental-noise inputs and ecological-variable 
outputs, AR(1) model, Clayton copula. Grey and black points give 
means over 50 replicate simulations of the listed statistics, 
grey points are for noise inputs and black points are for
model outputs. Error bars are standard errors. 
$P$-values (purple points, right axis) are for a 
paired $t$-test of the null hypothesis that the distributions have the same mean. 
For bottom panels, headers give Pearson correlations for
the points. The regression line through the points (solid) was similar
to the 1-1 line (dashed). 
See section \ref{Causes} for details.\label{fig_Cause4copula_C_kend_ar1}}
\end{center}
\end{figure}

@@

<!--adding plots showing influences of non-normal copula on spatial Taylors law-->
\begin{figure}[!h]
\textbf{ \hspace{2cm} (A) \hspace{2.92 cm} (B) \hspace{2.92 cm} (C) \hspace{2.92 cm} (D)} \\
\vspace{-0.8 cm} 
\begin{center}
\includegraphics[width=16 cm]{./Results/taylorslaw_results/SpatialTL_CNFSC_spearman.png}\\
\end{center}
\textbf{ \hspace{1 cm} (E) \hspace{1.4 cm} (F) \hspace{1.2 cm} (G) \hspace{1.2 cm} (H) \hspace{1.4 cm} (I) \hspace{1.4 cm} (J) \hspace{1.2 cm} (K)}\\
\vspace{-0.8 cm}
\begin{center}
\includegraphics[width=16 cm]{./Results/taylorslaw_results/SpatialTL_7stats_spearman.pdf}
\caption[Spatial Taylor's law results]{Spatial Taylor's law results. Log(spatial variance) vs. 
log(spatial mean) relationships for $1000$ simulations over $50$ time steps for $25$ locations 
using Clayton (A), normal (B), Frank (C), and survival Clayton (D) copula structures between
locations. Results from all $1000$ 
simulations were plotted on the same axes for each panel, but two example simulations are 
shown in solid and void black dots to help assess to what extent variation on the plots was between 
simulations or within simulations. For each simulation, we tested the linearity 
(E) and homoskedasticity (F) of the log(variance) vs. log(mean) plot for that simulation, 
and quantified the root mean squared error of points from the linear regression line (G).
We also calculated the linear regression intercept (H) and slope (I), the quadratic
term of a quadratic regression through the points on the log(variance) vs. log(mean) plot
(J), and the curvature of that quadratic regression (K). Distributions of values 
across all 1000 simulations are displayed. Spearman-preserving surrogates were used, 
though results using Kendall-preserving surrogates were similar. See section \ref{Consequences}
and Appendices \ref{SM-Consequence_TL_simulation} and 
\ref{SM-sect:copsurrognd} for details.\label{fig_spTL}}
\end{center}
\end{figure}
